Evaluation of Machine Translation in Languages

### 2024-07-13 paper/paper.tex
updated paper

@@

As of July 2024, 243 languages and languages varities are supported by Google Translate (according to Wikipedia \cite{wikipedia-google-translate})

Machine translation (MT) can be traced down to as early as 1949 \cite{weaver-1999}, recently growing as a popular research field within Natural Language Processing (NLP).

Statistical machine translation, which previously dominated MT research for many years with its reliance on various count-based models, has largely been surpassed by neural machine translation (NMT) \cite{stahlberg-2020-nmt-review}. Nowadays, Large Language Models (LLMs) such as mBart \cite{liu-2020-mbart}, M2M-100 \cite{fan-2020-m2m100}, NLLB-200 \cite{nllb200}, T5 \cite{raffel-2023-t5}, and GPT-4 \cite{openai-2024-gpt4-technical-report} dominated the field of machine translation. DeepL \cite{deepl} currently proclaimed itself as 'The world most accurate translator'.

@@

\section{Related Work}

As is the case of most fields within Natural Language Processing (NLP), the current state-of-the-art (SOTA) in machine translation (MT) is characterised by the significant advancements brought by large language models (LLMs) like OpenAI's GPT-4. These models have surpassed traditional neural machine translation (NMT) systems in several areas, offering more nuanced and accurate translations, especially in handling context and idiomatic expressions.

\section{Dataset}

\subsection{Overview}

@@

\subsection{Analysis}

@@

\section{Methodology}.

@@

>>>>

Machine translation (MT) is the task of automatically translating from one language to another. This can be done through text or audio. Ideally, machine translation should be able to conserve the contextual, idiomatic and pragmatic nuances of both languages.

Machine translation can be traced down to as early as 1949, \cite{weaver-1999}, with the first public demonstration of an MT system happened on 7th January 1954 through a collaboration with IBM \cite{hutchins-2006-first-mt}. A meticulously chosen set of 49 Russian sentences was translated into English, utilising a highly limited vocabulary of 250 words and only 6 grammar rules. While this had minimal scientific impact, it was impressive enough to encourage substantial funding for MT research in the USA.

The decade of high expectation and disillusion, 1956-1966 \cite{hutchins-2001-mt-50-years}.

The 1960s, particularly 1967-1976 is dubbed as 'the quiet decade' for machine translation by Hutchins \cite{hutchins-2001-mt-50-years}

Rule-based MT relied on extensive linguistic rules and bilingual dictionaries. They were based on syntactic and grammatical rules to parse and generate translations.

Since 1989, the dominance of the rule-based approach has been challenged by the rise of new methods and strategies, collectively referred to as ‘corpus-based’ methods \cite{hutchins-1994-research-methods-mt,hutchins-1998-development-mt}. Subsequently, statistics-based approaches for MT re-emerged, bolstered by the recent success of probabilistic techniques in
speech recognition.

1980s: Statistical Machine Translation (SMT) Emerges, use statistical models, probability

Late 2010s: Rise of Neural Networks
Neural Machine Translation (NMT): Starting around 2014, neural networks began to dominate MT. NMT systems, like those using the sequence-to-sequence (Seq2Seq) model with attention mechanisms (e.g., the Transformer model), demonstrated significant improvements in translation quality.
Advantages Over SMT: NMT models often produce more fluent and coherent translations as they consider the entire sentence context, rather than breaking it down into smaller units like phrases.

...recently growing as a popular research field within Natural Language Processing (NLP). Statistical machine translation \cite{lopez-2008-smt}, which previously dominated for many years in the early 2000s with its reliance on various count-based models, has largely been surpassed by neural machine translation (NMT) in (year). Since the introduction of Transformers in 2017 \cite{vaswani-2017-attention}, Natural Language Processing (NLP) and machine translation in particular reached a giant milestone. The following years saw the birth of Large Language Models (LLMs) such as BERT \cite{devlin-2019-bert} and GPT \cite{openai-2024-gpt4}.

Difference between LLMs and MT engines?
MT engines are specifically designed to translate text from one language to another while preserving the meaning and context as accurately as possible. MT engines can also be based on transformer architectures, but they are usually fine-tuned and optimized for the task of translation. They often incorporate additional components like alignment models and post-processing steps to ensure translation quality.

Primary Function: LLMs focus on generating and understanding natural language text, whereas MT engines focus exclusively on translating text between languages.
Training Data: LLMs are trained on diverse text data from various sources, while MT engines are trained specifically on parallel corpora (texts in two or more languages that are translations of each other).
Output: LLMs generate coherent text in the same language as the input, often in a more creative or conversational style. MT engines produce translated text in a different language, aiming for accuracy and fidelity to the original meaning.

@@

As of July 2024, 243 languages are supported by Google Translate (according to Wikipedia \cite{wikipedia-google-translate})

Many language models are trained on predominantly English NLP tasks, before fine-tuned for downstream tasks in other languages.

There is little insights on how the current state of machine translation is across different language. Therefore...

There does not seem to be a lot of pre-trained models that focuses on a single language. This could be because multilingual LLMs perform better? find proof

Focus on single-language models?

@@

\section{Literature Review}

DeepL \cite{deepl} is a machine translation service that has gained widespread acclaim, currently proclaimed itself as the world's most accurate translator. There is not much information on how the model or techniques as it is essentially a private business. (irrelevant?)

Instruction Learning and decoder-based models...

GPT \cite{openai-2024-gpt4}

Multilingual Language Models are on the rise for the task of machine translation that involves several languages

A major drawback within most multilingual LMs nowadays is that they are either trained, tested, prioritised on English first.

mBERT \cite{devlin-2019-bert} is the multilingual version of BERT, introduced in the same paper, trained on 104 different languages. However, the model been shown to suffer on low-resource languages \cite{wu-2020-mbert-are-all}.

M-BERT
does create multilingual representations,

T5 \cite{raffel-2023-t5} is a Text-to-Text Transfer Transformer model trained on English.

mt5 \cite{xue-2021-mt5} is a multilingual variant of T5 \cite{raffel-2023-t5}, covering a total of 101 languages.

PolyLM \cite{wei-2023-polylm}

mBart \cite{liu-2020-mbart}

NLLB \cite{nllb200}

MistralAI \cite{jiang-2023-mistral-7b}

Massive Parallel Corpora: The availability of large parallel corpora (datasets containing aligned text pairs in different languages) has been crucial for training effective translation models. Datasets like Europarl, OpenSubtitles, and Common Crawl provide a wealth of parallel text data.
Unsupervised Translation: Techniques have been developed to perform translation without the need for large parallel corpora, using monolingual data to train models that can translate between languages. This is especially useful for low-resource languages.

Multilingual and Zero-Shot Translation:

Multilingual NMT Models: Models like Google's multilingual NMT and Facebook's M2M-100 can translate directly between many language pairs without relying on an intermediate language (like English). These models are trained on datasets containing multiple languages simultaneously.
Zero-Shot Translation: These models can sometimes translate between language pairs they were not explicitly trained on, leveraging knowledge from related languages or shared linguistic features.

\section{Experiments}

\subsection{Dataset}

@@

@@

@@

GPT-4 \cite{openai-2024-gpt4} try GPT and see

READ THIS \cite{intento-2020} (exactly what this paper should do, maybe compare results and find insights)
